---
title: "HW1_Rodolfo_Urbano_ggplot2exercise"
author: "Rodolfo Urbano"
date: "Saturday, January 17, 2015"
output: html_document
---

##Reshape "iris" data frame & plot using ggplot2

```{r}
#Show the first few lines of the iris data frame.
head(iris)
```

##Data needs reshaping

"iris" is long and contains Length and Width measurements for plant parts (Petal & Sepal) as separate columns. Measurements will be merged and plant part information provided as a value in a new column.

```{r}
#load reshape 2 package
library(reshape2)
#create a new column that we will need to track row associations
iris$Flower_id <- rownames(iris)
head(iris) #you can now see the new column Flower_id
#melt data(wide to long) to condense the number of columns
iris_melted <- melt(iris)
```

##"iris" has been melted(Long -> Wide) into "iris_melted"
All measurements are now found in the "value" column with iris variables indicated in column "variables"

```{r}
head(iris_melted) 
```

## "variable" column indicating flower part and measurement type needs to be split into new columns. 

```{r}
#strsplit function splits "variable" as character in iris_melted at "."
split_variable <- strsplit(as.character(iris_melted$variable), split = "\\.")
#create two new columns and extract each variable that was split
iris_melted$flower_part <- sapply(split_variable, "[", 1)
iris_melted$measurement_type <- sapply(split_variable, "[", 2)
#remove the "variable" column
iris_melted$variable <- NULL
#use dcast(cast for data frame output) to reshape and separate measurement type
iris_cast <- dcast(iris_melted, formula = Flower_id + Species + flower_part ~ measurement_type)
```

## iris_cast is now in desired format

```{r}
head(iris_cast)
```

## Plot iris_cast Width & Length measurements using ggplot2
Facet by species and flower type


```{r}
#load ggplot2 package and plot data. 
library(ggplot2)
ggplot(data=iris_cast, aes(x=Width, y=Length, color = Species))+ # Add points and use free scales in the facet
    geom_point()+facet_grid(Species~flower_part, scale="free")+
    # Add a regression line
    geom_smooth(method="lm")+
    # Use the black/white theme and increase the font size
    theme_bw(base_size=12)
```